Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-05-09 ms/spaceness.tex
andy & cj overleaf edits

\author[$\ast$,1]{C.J. Battey}
\author[$\ast$]{Peter L. Ralph}
\author[$\ast$]{Andrew D. Kern}

@@

\begin{abstract}
    Real geography is continuous,
    but standard models in population genetics are based on discrete, well-mixed populations. 
    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, 
    but are applied to real populations that have significant geographic structure.
    Here we use simulations of populations living in continuous geography 
    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
    We find that several common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered. 
    Standard estimates of population size history based on the site frequency spectrum become biased with lower dispersal. 
    We also show that the combination of spatially autocorrelated environments and limited dispersal
    can cause genome-wide association studies to identify numerous spurious signals of genetic association 
    with purely environmentally determined phenotypes,
    and that this bias is only partially corrected by regressing out principal components of ancestry. 
    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

\correspondingauthoraffiliation{1}{301 Pacific Hall, University of Oregon Dept. Biology, Institute for Ecology and Evolution. cbattey2@uoregon.edu.}
\vspace{-35pt}% Only used for adjusting extra space in the left column of the first page

\plr{note: address confounding of $\sigma$ and neighborhood size in disco}

\plr{note: what are conclusions about bias or not of Ne(t)?}

@@

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. 
However, mates tend to live near to one another and to their offspring,
leading to a positive correlation between genetic and geographic distances. 
% Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to live nearby to both their parents, in particular when dispersal and mate choice distances are short. 
This pattern of ``isolation by distance'' \citep{Wright1943}, describing mean relatedness as a function of distance in flat landscapes,
is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
(e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}),
much modern work still describes geographic structure as a set of populations connected by migration 
\citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}.
For this reason,
most population genetics statistics are interpreted with reference to discrete, well-mixed populations,
and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000},
effectively assuming these are randomly mating units.

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. 
Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}.
The key assumption is actually that the individuals of each generation are \emph{exchangeable},
so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}.
If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways.
For instance, if mean viability or fecundity is spatially autocorrelated,
then limited geographic dispersal will lead to parent--offspring correlations.
Futhermore, nearby individuals will be more closely related than an average random pair,
so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population. 
It has long been appreciated that this model misspecification subjects downstream inferences to bias, 
but the extent and nature of these effects remain largely uninvestigated.

For instance, nonzero values of Tajima's $D$ are often interpreted as reflecting evidence of selection or past population size changes \citep{tajima1989statistical}.
This statistic is a summary of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals.
Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972},
which can affect the value of Tajima's $D$.
% Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population 
% these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations. 
Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only
historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space,
both to evaluate the assumptions underlying existing methods, but also to identify where the most promising signals of geography lie.

A related issue is the spatial distribution of sampling effort --
since nearby individuals are more likely to be closely related than distant ones,
observed patterns of relatedness are expected to depend on the geographic sampling scheme,
perhaps strongly.
In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, 
but the effects of this process on many downstream inference procedures is unknown.
\cjb{must be more out there on sampling -- any recs for papers to read+cite?}

The issue of fine-scale geographic structure may have particularly important implications for genome-wide association studies (GWAS). 
This is because many phenotypes of interest have strong geographic differences due to the (nongenetic) influence of environmental or socioeconomic factors,
which can therefore show spurious correlations with spatially patterned allele frequencies \citep{Bulik-Sullivan2015,Mathieson2012}. 
This may be particularly important for the study of selection on polygenic traits,
whose heritable genetic components are determined by many loci of weak effect,
those hardest to disentangle from spurious correlations. 
Indeed, two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient. However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simluate chromosome-scale genomic data in tens of thousands of individuals, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Lastly, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation Wright model is that this balance is nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual.
\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population 
approaches the random mating expectation when $\rho\sigma^2 > 1$, 
and proposed that this had the important implication that most population genetic expectations for randomly mating populations 
could be applied to continuously distributed populations with relatively little error.

The first approach to modeling continuously-distributed populations
was to endow individuals in a Wright-Fisher model with locations in continuous space.
However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
which (aside from being implausible) \cite{Felsenstein1975} showed would eventually lead to unrealistic population clumping
if the species range is sufficiently large.
After global population regulation,
the next method for modeling spatial populations might to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
thus enforcing regular population density by edict.
Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
\cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
Although these grid models are likely good approximations of continuous geography in many situations,
these do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
assumptions whose impacts are unknown. 
An alternative method for dealing with continuous geography is a new class of coalescent models,
the Spatial Lambda Fleming-Viot models \citep{Barton2010,Kelleher2014}.

To avoid questionable assumptions,
we worked with forwards-time, individual-based simulations.
By scaling the probability of survival in each timestep to local population density, 
we shift reproductive output towards regions of low-density, which prevents populations from clustering. 
% LOTS of people have done models like this: classic ecology?
% A similar approach was taken previously by \citet{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients. 
Such models have been used extensively in ecological modeling but rarely in population genetics,
where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, 
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations 
and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We simulated populations using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is equal to the mean parent-offspring distance.  
A key parameter we report below is
\citeauthor{Wright1943} citeyearpar{Wright1943} ``neighborhood size'', 
defined to be $N_W = 4\pi\sigma^2 \rho$ where $\rho$ is the population density.
This the approximate number of individuals available for mating in our simulation.

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ 
who disperse a random, Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates, reflected to stay within the species range.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$ 
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size $10^{8}$, 
recombination rate $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
Simulations were run for 1.6 million timesteps (approximately $30N$ generations), or until all extant individuals shared a common ancestor within the simulation (i.e., the tree sequence had coalesced). 
\ak{maybe worth including a table with some basic runtime results in the supplement?}

@@

Our demographic model includes paramters for population density ($K$),
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations. 
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

@@

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples. The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{chapman2002effect}.

@@

%%%%%%%%%%%
\subsection{Demographic Modeling}

We fit single-population demographic models to the site frequency spectra of 20 individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}. This analysis was replicated across random, point, and midpoint sampling strategies. Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}, 
and 100 bootstrap replicates were generated for each simulation by resampling over sites. 
\plr{what were bootstrap replicates used for? if anything, need to say more precisely what you mean here}
Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $N_e$ per generation was used to represent the output of each simulation.

In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models. All coalescent simulations were performed using msprime \citep{Kelleher2016}. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

@@

To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with principal components as covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero and standard deviation 10
to a mean that may depend on spatial position.
We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} phenotypes, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype form 90 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $110 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 130 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 130.

>>>>

\author[$\ast$,1]{C.J. Battey}
\author[$\ast, \dagger$]{Peter L. Ralph}
\author[$\ast, \dagger$]{Andrew D. Kern}

@@

\begin{abstract}
    Real geography is continuous,
    but standard models in population genetics are based on discrete, well-mixed populations. 
    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, but are applied to clustered samples from populations that are structured clinally over space.
    Here we use simulations of populations living in continuous geography 
    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
    We find that several common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered.
    The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance. 
    We also show that the combination of spatially autocorrelated environments and limited dispersal
    can cause genome-wide association studies to identify numerous spurious signals of genetic association 
    with purely environmentally determined phenotypes,
    and that this bias is only partially corrected by regressing out principal components of ancestry. 
    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

\correspondingauthoraffiliation{1}{301 Pacific Hall, University of Oregon Dept. Biology, Institute for Ecology and Evolution. cbattey2@uoregon.edu.}
\correspondingauthoraffiliation{$\dagger$}{these authors co-supervised this project}
\vspace{-35pt}% Only used for adjusting extra space in the left column of the first page

@@

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}. Despite a long history of analytical work describing the genetics of populations distributed across continuous geography (e.g., \cite{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation}), much modern work still describes geographic structure as a set of populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations}. For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}. The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}. If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways. For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations. Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population. It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.

Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection, the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}). Inferring a trajectory of population sizes through time has therefore become a major goal in population genetics, with many recent methods proposed to accomplish this task \citep{Liu2015,Terhorst2016,Schiffels2014,Sheehan2013}. All of these methods assume that the populations being modeled are approximately randomly mating (though this also depends on how researchers interpret the inferred parameter $\Theta=4N\mu$). Because isolation by distance is expected to change the rate of drift in a population \citep{Maruyama1972}, it may cause biases in inferred population sizes. However the effects of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.

Spatial structure is also a major challenge for interpreting the results of genome-wide association studies (GWAS). This is because many phenotypes of interest have strong geographic differences due to the (nongenetic) influence of environmental or socioeconomic factors,
which can therefore show spurious correlations with spatially patterned allele frequencies \citep{Bulik-Sullivan2015,Mathieson2012}. Indeed, two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient. However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS.

Last, some of the most basic tools of population genetics are summary statistics like $F_{is}$ and Tajima's $D$, which are often interpreted as reflecting the influence of selection or demography on sampled populations \citep{tajima1989statistical}. Statistics like Tajima's $D$ are essentially summaries of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals. Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972}, which can affect the value of Tajima's $D$. Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space, both to evaluate the assumptions underlying existing methods and to identify where the most promising signals of geography lie.

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and Drosophila, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Last, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
is determined by the interaction between rates of migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation Wright model is that this balance is nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (\ref{table:NStable}), and \cite{Wright1946} estimated values of 500-1000 for \textit{Drosophila pseudoobscura}.
\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population 
approaches the random mating expectation when $\rho\sigma^2 > 1$, 
and proposed that this had the important implication that most population genetic expectations for randomly mating populations 
could be applied to continuously distributed populations with relatively little error.

The first approach to modeling continuously-distributed populations
was to endow individuals in a Wright-Fisher model with locations in continuous space.
However, since the total size of the population is constrained, this introduces interactions between arbitrarily distant individuals,
which (aside from being implausible) \cite{Felsenstein1975} showed would eventually lead to unrealistic population clumping
if the range is sufficiently large.
Another method for modeling spatial populations is to assume the existence of a grid of discrete randomly-mating populations connected by migration, 
thus enforcing regular population density by edict.
Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, 
\cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. 
Although these grid models are likely good approximations of continuous geography in many situations,
they do not model demographic fluctuations, and limit investigation of spatial structure below the level of the deme,
assumptions whose impacts are unknown. 
An alternative method for dealing with continuous geography is a new class of coalescent models,
the Spatial Lambda Fleming-Viot models \citep{Barton2010,Kelleher2014}.

To avoid questionable assumptions, we here used forwards-time, individual-based simulations. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering. Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We simulated populations using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is equal to the mean parent-offspring distance.

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
who disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates, reflected to stay within the species range.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$ 
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

Our demographic model includes parameters for population density ($K$),
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations. 
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

@@

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples. The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}. For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (>48 map units) and nearby (<2 map units) individuals. Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance among individuals carrying each derived allele in a set of simulations representing a range of neighborhood sizes.

@@

%%%%%%%%%%%
\subsection{Demographic Modeling}
To assess the impacts of continuous spatial structure on demographic inference we inferred population size histories for all simulations using two approaches: Stairwayplot \citep{Liu2015} and SMC++ \citep{Terhorst2016}. Stairwayplot fits its model to a genome-wide estimate of the SFS, while SMC++ also incorporates information about linkage disequilibrium across the genome in a sequentially markovian coalescent framework. For both methods we sampled 20 individuals from all spatial simulations using random, midpoint, and point sampling strategies.

Stairwayplot documentation suggests fitting models across multiple bootstrap replicates drawn from empirical genomic data and taking the median inferred $N_{e}$ per unit time as the best estimate. We calculated site frequency spectra in scikit-allel \citep{Miles2017}, generated 100 bootstrap replicates per simulation by resampling over sites, and fit models for all bootstrap samples using default settings in stairwayplot.

For SMC++ we first converted simulated genotype matrices to VCFs with msprime and then used SMC++'s standard pipeline for preparing input files assuming no polarization error in the SFS. We used the first individual in the VCF as the "designated individual" when fitting models, and allowed the program to estimate the recombination rate during optimiziation. We fit models using the 'estimate' command rather than the now recommended cross-validation approach because our simulations had only a single contig.

To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy, and compared these values to the mean $N_{e}$ of random-mating simulations estimated from $\Theta_{\pi}$. We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model and plotting these against neighborhood size.

%In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models. All coalescent simulations were performed using msprime \citep{Kelleher2016}. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

@@

To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with principal components as covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe \ak{need citation}. Conceptually our approach is similar to that taken in \cite{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero and standard deviation 10
to a mean that may depend on spatial position.
We adjust the geographic pattern of mean phenotype to reflect spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

We performed principal components analysis (PCA) using scikit-allel \citep{Miles2017} on the matrix of derived allele counts by individual for each simulation. 
SNPs were first filtered to remove strongly linked sites by calculating LD between all pairs of SNPs in a 200-SNP moving window and dropping one of each pair of sites with an $R^2$ over 0.1. The LD-pruned allele count matrix was then centered and all sites scaled to unit variance when conducting the PCA, following recommendations in \cite{Patterson2006}.
